The Galactic Chemical Evolution of Carbon

### 2023-05-01 thesis/tex/main.tex
incramentally more cites and glossaries

%! tex program = lualatex
\DocumentMetadata{testphase=phase-III, lang=en-US}

@@

\tagpdfsetup
{
    activate-all,
}

\usepackage{amsmath}	% Advanced maths commands

\usepackage[T1]{fontenc}
\usepackage{txfonts}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=magenta,
    filecolor=black,
    urlcolor=blue,
    citecolor=blue,
    pdftitle={The Galactic Chemical Evolution of Carbon},
}

@@

\usepackage{aas_macros}
\usepackage{natbib}
\usepackage[nottoc,numbib]{tocbibind}

@@

% these don't work with lualatex/phase-iii
%\usepackage{accsupp}
%\usepackage[accsupp]{axessibility}
% \pdfcompresslevel=0

% accessibility
% \usepackage{tagpdf}

% \usepackage[tagpdf]{axessibility}

%%%%%%%%%%%%%%% Acronyms %%%%%%%%%%%%%%%%%%
\usepackage[acronym,automake,symbols,nogroupskip,counter=section]{glossaries-extra}
\newglossary*{models}{Stellar Yield Models}
\makeglossaries

\setabbreviationstyle[acronym]{long-short-sc}

% \newabbreviation
% [description={massive star explosions}]
% {cc}{ccsne}{core collapse supernovae}

\newacronym
[ description={heelloo}, seealso={ccg} ]
{cc}{ccsne}{core collapse supernovae}

@@

\newacronym
{dla}{dla}{damped Lyman-alpha systems}

@@

@@

\glsxtrnewsymbol
[description = {The symbol for the sun}]
{sun}{\ensuremath{\odot}}

@@

\newcommand{\et}{\gls{eta}}
\newcommand{\zet}{\gls{zeta}}
\newcommand{\sfr}{\gls{sfr}}

\newglossaryentry{metallicity}{name={metallicity},
    description={the (mass) fraction of a star or gass which is not made of either hydrogen or helium. For the sun, the metallicity is $\Zo = 0.014$}
}

@@

\newglossaryentry{ccg}
{
    name={core collapse supernovae},
    description ={massive star explosions. core collapse supernovae (\textcs{ccsne}) produce many elements including $\alpha$-elements, Fe-peak elements, and r and s-process elements.}
}

\newglossaryentry{c11}
{
    name={C11},
    type={models},
    description={the yield set from \citet{cristallo+11,cristallo+15}. The models were run at $m= (1.3, 1.5, 2.0, 2.5, 3.0, 4.0, 5.0, 6.0)$ and $Z=(0.0001, 0.0003, 0.001, 0.002, 0.003, 0.006, 0.008, 0.01, 0.014, 0.02)$. },
}
\newglossaryentry{k10}
{
    name={K10},
    type={models},
    description={the yield set from \citet{karakas10}. Includes models run at 
    $M=(1.0, 1.25, 1.5, 1.75, 1.9, 2.25, 2.5, 3.0, 3.5, 4.0, 4.5, 5.0, 5.5, 6.0)$ and $Z=(0.0001, 0.004, 0.008, 0.02)$},
}
\newglossaryentry{v13}
{
    name={V13},
    type={models},
    description={the \agb\ yield set from \citet{ventura+13,ventura+14,ventura+18,ventura+20}. Includes models with masses $m=(1.5, 2.0, 2.5, 3.0, 3.5, 4.0, 4.5, 5.0, 6.0, 6.5, 7.0)$ and with metallicities $Z=(0.0003, 0.001, 0.002, 0.004, 0.008, 0.014, 0.04)$},
}
\newglossaryentry{k16}
{
    name={K16},
    type={models},
    description={The yield set from \citet{KL16, karakas+18}. The model is sampled at 
        $M=(0.0003, 0.001, 0.002, 0.004, 0.008, 0.014, 0.04)$
        and $Z=(0.0003, 0.001, 0.002, 0.004, 0.008, 0.014, 0.04)$
    },
}

\newcommand{\cxi}{\gls{c11}}
\newcommand{\kx}{\gls{k10}}
\newcommand{\kxvi}{\gls{k16}}
\newcommand{\vviii}{\gls{v13}}

\newcommand{\Ycc}{\gls{ycc}}
\newcommand{\Yoc}{\gls{ymg}}
\newcommand{\Ycagb}{\gls{ycagb}}
\newcommand{\sun}{\gls{sun}}
\newcommand{\Z}{\gls{Z}}
\newcommand{\Zo}{\gls{Z}_{\gls{sun}}}

@@

%%%%%%%%%%%%%%%%%%% Citation commands %%%%%%%%%%%%%%%%%
\defcitealias{cristallo+11}{C11}
\defcitealias{cristallo+15}{C15}
\defcitealias{ventura+13}{V13}
\defcitealias{karakas10}{K10}
\defcitealias{KL16}{KL16}
\defcitealias{karakas+18}{K18}
\defcitealias{james+21}{J21}

\newcommand{\JJ}{\citetalias{james+21}}
\newcommand{\cristallo}{\citetalias{cristallo+11}+\citetalias{cristallo+15}}
\newcommand{\karakas}{\citetalias{karakas10}}
\newcommand{\kl}{\citetalias{KL16}+\citetalias{karakas+18}}
\newcommand{\ventura}{\citetalias{ventura+13}}

\newcommand{\VICE}{\texttt{VICE}}
\newcommand{\caah}{[C/Mg]-[Mg/H]}
\newcommand{\caafe}{[C/Mg]-[Mg/Fe]}

\newcommand\T{\rule{0pt}{2.6ex}}       % Top strut
\newcommand\B{\rule[-1.2ex]{0pt}{0pt}} % Bottom strut
\newcommand{\possesivecite}[1]{\citeauthor{#1}'s \citeyearpar{#1}}
\newcommand{\alp}{$\alpha$}

@@

\setcitestyle{aysep={}} 
\setcitestyle{notesep={; }}

\title{Carbon Nucleosynthesis}
\author{Daniel A. Boyea}

@@

\begin{document}

\linespread{1.25}

@@

\begin{titlepage}
   \begin{center}
       \textbf{The Galactic Chemical Evolution of Carbon}\\
       Implications for Stellar Nucleosynthesis\\
       \vspace*{3\baselineskip}
        Undergraduate Research Thesis\\
       \vspace*{3\baselineskip}
    Presented in partial fulfillment of the requirements for graduation \textit{with research distinction in Astronomy and Astrophysics} in the College of Arts and Sciences of The Ohio State University\\
       \vspace*{3\baselineskip}
        by\\
       \vspace*{3\baselineskip}
       {Daniel Alexander Boyea\\
       \vspace*{3\baselineskip}
       The Ohio State University\\
       April 2023\\
       \vspace*{3\baselineskip}
       Project Advisors\\
       Professor David H. Weinberg, Department of Astronomy \\
       James W. Johnson, Department of Astronomy
       \vfill
   \end{center}
\end{titlepage}

@@

Even though you may have not been able to help me with the science, dear friends and family, you have been no less importaint in the completion of this work. 
Eric, even across the country, you have always been a point of support and stability, and I am always happy to be around you. Anya, you are a wonderful friend and I cannot thank you enough for your unconditional support this year, 
All my friends who are just wonderful, inspiring, and incredible people---Kaia, Maria, Autumn, Simon, Bary, Aaliyah, Sanskruti, Alyssa, Harrison, Denis, Keith, and so many others. 
And my family (including Arya, my most lovable doodle)---I can not eveen being to thank you enough for simply helping me recover my health and make it to graduation.

@@

\Gce{} aims to understand the chemical and star formation histories of galaxies. At the core of \gce{} models are stellar yields---the amount of each chemical element stars produce (or destroy). Each element is produced in different amounts by different stars, leaving traces of the Galaxy's evolutionary history. Through \textit{Galactic Archeology}, we can reconstruct the history of our galaxy by investigating clues left behind in stars.

Here, I aim to understand C enrichment---where it is produced and how its abundances evolve over time. C is unique nucleosynthetically, one of few light elements (along with N) produced in \agb{} stars \citep[e.g.][]{jennifer19, KL14}. C and N are well-studied elements since they are easy to observe in stellar spectra, even at the lowest metallicities \cite[e.g.][]{fabbian+09, nissen+14, lambert81, laird85, lambert86}. Additionally, C and N are used as age indicators in red giant branch stars \citep{martig16, MG15, hasselquist19, vincenzo+21}.

Gas-phase [C/O] ratios%
%
\footnote{In this paper, I use the standard notation for chemical abundances. $[A/B] = \log_{10}\left(A/B\right) - \log_{10}\left(A_{\sun}/B_{\sun}\right)$, i.e. $[A/B]$ is the logarithm of the ratio between A and B, scales such that $[A/B]=0$ for the sun. Solar abundances are as measured in \citet{asplund+09}}%
%
---observed in very low \gls{metallicity}, high redshift \dla---decrease with increasing [O/H] \citep{FN15, cooke+17}. Then, [C/O] increases above $\rm [O/H] \approx -1$ (\citealt{berg+19}; see discussion in \S\ref{sec:gas}).
While we know C is produced in \agb{} stars and \cc{}, we still have a limited understanding of the magnitude and metallicity-dependence of each process.

@@

Accurate determination of stellar birth abundances for C and N is challenging. When a star enters the red giant branch, material from the CNO-processed core is mixed with the envelope in first dredge up, enhancing N and depleting C \citep{iben67, vincenzo+21,KL14}. Measurements of these evolved stars'  atmospheres will no longer reflect their birth abundances.  Additionally, gas-phase measurements of C are extremely limited as C lacks strong lines in HII regions \citep{skillman+20}.

To this end, I use a sample of subgiant stars from the \apogee{} \citep{apogee17}. According to stellar evolution theory and observations \citep{gilroy89, korn+07, lind+08, souto+18, souto19}, these stars have not yet experienced first dredge up but have well-mixed photospheres. Their photosphoric abundances should thus represent their birth composition.  In Fig.~\ref{fig:subgiants}, I plot this sample, selected by the criteria in Roberts et al. (2023, in prep; see Appendix \ref{sec:jack}). These empirical results contain information on the relative C and Mg yields, which I use later in this work.

\begin{figure}[htp]
    \includegraphics[width=\textwidth]{subgiants_mgh.pdf}
    \includegraphics[width=\textwidth]{subgiants_mgfe.pdf}
    \caption[APOGEE Subgiants]{The [C/Mg] ratio against [Mg/H] (top) and [Mg/Fe] (bottom) for the \citetjack~sample of \apogee{} subgiants. On the top, I plot high and low-$\alpha$ stars in blue and orange, using the separation defined in Equation \ref{eq:high_alpha}. On the bottom, I color-code stars according to their [Mg/H] abundance.}
    \label{fig:subgiants}
\end{figure}

@@

Theoretical models of stellar nucleosynthesis provide the starting point of this investigation. I focus on three primary nucleosynthetic pathways: \agb{} stars, \cc{}, and \ia. Each process has unique timescales and yields, traceable through the tools of \gce. I compare C, produced in \cc{} and \agb\ stars, to Mg and Fe as tracers of \cc{} and \ia{} espectively.

@@

For an element X and star with mass $M$, the stellar yield $\tilde{y}$ is defined as the net production of X relative to $M$, or
\begin{equation}
    \tilde{y}_{\rm X} = \frac{M_{\rm X,\ ejected} - Z_{0, X} M_{\rm ejected}}{M}   
\end{equation}
where $M_{\rm ejected}$ and $M_{\rm X, ejected}$  are the total ejected mass of the envelope and the element X, respectively. For example, if a 1 $M_\odot$ star has $\tilde{y}_{\rm C} = 10^{-3}$, then the star will add $10^{-3}\ M_\odot$ of C to the interstellar medium. 
Although per-star yields are necessary to compute \agb{} star enrichment rates in \gce{}  models, \imf-averaged yields are useful in interpreting their predictions. For a yield $y$ from a star of mass M and initial metallicity $Z$, the \imf-averaged yield is given by 
\begin{equation} \label{eq:imf-yield}
    y_{\rm X}(Z) = 
    \int_{M_{\rm min}}^{M_{\rm max}} 
    \tilde{y}_{\rm X}(M, Z)
    \frac{dN}{dM}  \ dM
\end{equation}
where ${dN}/{dM}$ is the normalized \imf, and $M_{\rm min}$ and $M_{\rm max}$ are the minimum and maximum mass of stars, which I take to be $0.08 M_{\sun}$ and $100 M_{\sun}$ respectively. For \agb{} stars, the yields are truncated above 8 $M_{\odot}$. 
To calculate the \imf-weighted net yields, I use the Versatile Integrator for Chemical Evolution code (\VICE\footnote{\VICE~is available at \url{https://github.com/giganano/VICE}}).

@@

\begin{table}
	\centering
    \caption[Fiducial Model]{Yields for the fiducial model in units of \ssp~birth mass. See \S\ref{sec:agb} for the definition of \cxi.}
	\label{tab:fiducial_mod}

@@

@@

An \agb\ star is a low mass ($\lesssim 8 M_{\sun}$) star during its final phase of evolution.  In an \agb\ star, two competing processes determine the outcome of C production: third dredge up and hot bottom burning.  Third dredge up accompanies thermal pulses in \agb\ stars. During a third dredge up event, material from the CO core is mixed with material exposed to partial He-shell burning, increasing C and O abundances in the envelope \citep{KL14}. If this envelope is lost during the \agb\ phase, then C yields are enhanced.
Hot bottom burning is the activation of proton capture reactions, and the CNO cycle%
\footnote{The CNO (I) cycle is
\isotope[12]{C}(p, $\gamma$)
\isotope[13]{N}($\beta^+ \nu_e$)
\isotope[13]{C}(p, $\gamma$)
\isotope[14]{N}(p, $\gamma$)
\isotope[15]{O}($\beta^{+}\nu_e$)
\isotope[15]{N}(p, \alp)
\isotope[12]{C}.
 Other minor branches of the CNO cycle are of minor importance to C yields
 \citep{solar-fusion}.
}
at the base of the convective envelope. Because the $^{14}$N(p, $\gamma$) proton capture is the slowest component of the CNO cycle \citep{solar-fusion}, the CNO cycle converts nearly all \isotope[12]{C} into \isotope[14]{N}.
As a result, when both third dredge up and Hot bottom burning occur, \isotope[12]{C} yields are lowered (see discussion in \citealt{james+23} and \citealt{ventura+13}).

In this work, I explore four different sets of \agb\ star yield tables from literature, providing well-sampled grids in metallicity and mass for use in chemical evolution models. 
\begin{itemize}
    \item \cxi \citet{cristallo+11, cristallo+15}
    \item K10: \citet{karakas10}
    \item V13: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item K16: \citet{KL16} \citet{karakas+18}
\end{itemize}
I use \cxi table, amplified by a factor of 2.9, as the fiducial \agb\ yield.

Fig.~\ref{fig:y_agb} compares the \agb\ C yield $\tilde{y}_\text{C}^\text{AGB}(M, Z)$ for each model I consider. 
The yields may be negative if a star ejects material with a lower average C abundance than the material the star was formed from.
Most models agree on the qualitative shape of the net fractional \agb\ C yield---% 
stars with masses between \about{2~M_\odot} and \about{4~M_\odot} have the highest fractional C yields, with the mass of the peak increasing and overall yields decreasing with increasing $Z$.  High mass, high $Z$ stars destroy \isotope[12]{C} because they experience both third dredge up and hot bottom burning, but the latter is much more efficient.

In Fig.~\ref{fig:agb-ssp}, I plot the total production of C by \agb\ stars in a \ssp{} at an age $t$. 
Effectively, this is equivalent to integrating Eq.~\ref{eq:imf-yield} down to $M_{to}$, the turnoff mass for time $t$.
({In our model, the mass-lifetime relation is
$\log \tau_M = 1.02 - 3.57\log M + 0.90 \left(\log M\right)^2$,
where $\tau_M$ is in Gyr,
from \citealt{larson74}.})
As the mass range $2\lesssim M \lesssim 4$ is most important for C production, about half of C production occurs before \about{1} Gyr, similar to \ia\ Fe. 
K10 and K16 weight C production more heavily towards massive stars resulting in a faster enrichment delay time, whereas the \cxi\ and V13 models predict a slightly longer timescale of \about{1} Gyr, but little to no C is produced more than 2 Gyr after a star formation event. This is in contrast to Fe whos production lasts up to 10 Gyr after a star formation event. As shown in the right panel of Fig.~\ref{fig:agb-ssp}, with increasing $Z$, C enrichment occurs earlier, and C destruction in low-mass stars leads to a declining C abundance at late times.

\begin{figure}
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[Low mass C yields]{The net fractional \agb\ C yield  plotted as a function of initial stellar mass $M$ and color-coded according to metallicity. The black dashed line shows $\tilde{y}=0$ for reference. Each panel represents yields from one of four \agb\ models: \cxi, K10, V13, K16 (see \S \ref{sec:agb}). }
        \label{fig:y_agb}
\end{figure}
\begin{figure}
    \includegraphics[scale=1]{y_agb_t2.pdf}

>>>>

\DocumentMetadata{lang=en-US}

@@

\usepackage{amsmath}
\usepackage{siunitx}

\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{microtype}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=black,
    filecolor=black,
    urlcolor=black,
    citecolor=black,
    pdftitle={The Galactic Chemical Evolution of Carbon},
    pdfauthor={Daniel A Boyea},
    pdfcreator={Themself},
    pdfkeywords={astronomy,supernovae, stellar yields, chemical evolution, galaxies, carbon},
}

@@

\usepackage{aas_macros}
\usepackage{natbib}
\usepackage[nottoc,numbib]{tocbibind}
\setcitestyle{aysep={}} 
\setcitestyle{notesep={; }}

@@

\newcommand\T{\rule{0pt}{2.6ex}}       % Top strut
\newcommand\B{\rule[-1.2ex]{0pt}{0pt}} % Bottom strut

\usepackage{accsupp}
\usepackage[accsupp]{axessibility}

%%%%%%%%%%%%%%% Acronyms %%%%%%%%%%%%%%%%%%
\usepackage[acronym,automake,symbols,
nogroupskip,
accsupp,
counter=section,
postpunc={.},
]{glossaries-extra}

\newglossary*{models}{Stellar Yield Models}
\makeglossaries

\setabbreviationstyle[acronym]{long-short-sc}
\renewcommand*{\glsentrycounterlabel}{\theglossaryentry:\space}

\newacronym
[ description ={Core collapse supernovae. Massive star explosions. {\sc ccsne} produce many elements including $\alpha$-elements, Fe-peak elements, and r and s-process elements.}
]
{cc}{ccsne}{core collapse supernovae}

@@

\newacronym
[
description={Damped Lyman-alpha system. {\sc dla}s are clouds of gas from the 
early universe which are observed through their absorption of quasar spectra. The name 
comes from the strong Lyman-alpha lines ($\qty{1216}{\angstrom}$) due to H absorption.}
]
{dla}{dla}{damped Lyman-alpha systems}

@@

%%%%%%%%%%%%%%%%%%% Symbols %%%%%%%%%%%%%%%%%%%%%%%%
\glsxtrnewsymbol
[description = {The abundance ratio between elements $A$ and $B$. See }]
{abundratio}{\ensuremath{[A/B]}}

\glsxtrnewsymbol
[description = {The mass outflow loading factor}]
{aagb}{\ensuremath{\alpha_{\rm CC}}}

\glsxtrnewsymbol
[description = {The mass outflow loading factor}]
{acc}{\ensuremath{\alpha_{\rm CC}}}

\glsxtrnewsymbol
[description = {The massive star carbon yield metallicity dependence}]
{sfd}{\ensuremath{\dot{\Sigma}}}

\glsxtrnewsymbol
[description = {The mass of the sun. $M_\odot = \qty{1.989e33}{g}$}]
{msun}{\ensuremath{M_\odot}}

\glsxtrnewsymbol
[description = {The fractional contribution of AGB stars to the total \gls{imfave} carbon yield at solar metallicity.}]
{fagb}{\ensuremath{f_{\rm AGB}}}

\glsxtrnewsymbol
[description = {The net fraction initial-mass-function weighted low-mass star yield of carbon}]
{ytcagb}{\ensuremath{\tilde{y}_{\rm C}^{\rm AGB}}}

@@

\glsxtrnewsymbol
[description = {The metallicity for the sun}]
{zsun}{\ensuremath{Z_\odot}}

@@

%%%%%%%%%%%%%%%%%%%%%% terms %%%%%%%%%%%%%%%
\newglossaryentry{metallicity}{name={metallicity},
    description={the (mass) fraction of a star or gass which is not made of either hydrogen or helium. For the sun, the metallicity is $\Zo = 0.014$}
}

\newglossaryentry{nucleosynthesis}{name={nucleosynthesis},
    description={the synthesis of new elements inside stars}
}

\newglossaryentry{yield}{name={yield},
    description={the synthesis of new elements inside stars}
}

\newglossaryentry{multizone}{name={multi-zone},
    description={the synthesis of new elements inside stars}
}

\newglossaryentry{onezone}{name={one-zone},
    description={the synthesis of new elements inside stars}
}

@@

\newglossaryentry{rgb}{name={red giant branch},
    description={Red giant branch stars are stars which have completed hydrogen core burning and have expanded in size.}
}

\newglossaryentry{tdu}{name={third dredge up},
    description={Third dredge up occurs when ....}
}

\newglossaryentry{fdu}{name={first dredge up},
    description={Third dredge up occurs when ....}
}

\newglossaryentry{hbb}{name={hot bottom burning},
    description={Third dredge up occurs when ....}
}

\newglossaryentry{subgiant}{name={subgiant},
    description={A subgiant is...}
}

\newglossaryentry{imfave}{name={\gls{imf}-averaged},
    description={Initial-mass function ....}
}

%%%%%%%%%%%%%%%%%%%%%%%%%% Models %%%%%%%%%%%%%%%%%%%%%%%
\newglossaryentry{c11}
{
    name={C11},
    type={models},
    description={the yield set from \citet{cristallo+11,cristallo+15}. The models were run at $M = $ 1.3, 1.5, 2.0, 2.5, 3.0, 4.0, 5.0, 6.0 and $Z= $ 0.0001, 0.0003, 0.001, 0.002, 0.003, 0.006, 0.008, 0.01, 0.014, 0.02. },
}
\newglossaryentry{k10}
{
    name={K10},
    type={models},
    description={the yield set from \citet{karakas10}. Includes models run at 
    $M= $ 1.0, 1.25, 1.5, 1.75, 1.9, 2.25, 2.5, 3.0, 3.5, 4.0, 4.5, 5.0, 5.5, 6.0 and $Z=$ 0.0001, 0.004, 0.008, 0.02},
}
\newglossaryentry{v13}
{
    name={V13},
    type={models},
    description={the \agb\ yield set from \citet{ventura+13,ventura+14,ventura+18,ventura+20}. Includes models with masses $m=(1.5, 2.0, 2.5, 3.0, 3.5, 4.0, 4.5, 5.0, 6.0, 6.5, 7.0)$ and with metallicities $Z=(0.0003, 0.001, 0.002, 0.004, 0.008, 0.014, 0.04)$},
}
\newglossaryentry{k16}
{
    name={K16},
    type={models},
    description={The yield set from \citet{KL16, karakas+18}. The model is sampled at 
        $M=(0.0003, 0.001, 0.002, 0.004, 0.008, 0.014, 0.04)$
        and $Z=(0.0003, 0.001, 0.002, 0.004, 0.008, 0.014, 0.04)$
    },
}

\newcommand{\cxi}{\texttt{\gls{c11}}}
\newcommand{\kx}{\texttt{\gls{k10}}}
\newcommand{\kxvi}{\texttt{\gls{k16}}}
\newcommand{\vxiii}{\texttt{\gls{v13}}}

\newcommand{\Ycc}{\gls{ycc}}
\newcommand{\Yoc}{\gls{ymg}}
\newcommand{\Ycagb}{\gls{ycagb}}
\newcommand{\sun}{\odot}
\newcommand{\Z}{Z}
\newcommand{\Zo}{Z_{\sun}}

@@

\defcitealias{james+21}{J21}

\newcommand{\JJ}{\citetalias{james+21}}
\newcommand{\VICE}{\texttt{VICE}}
\newcommand{\caah}{[C/Mg]-[Mg/H]}
\newcommand{\caafe}{[C/Mg]-[Mg/Fe]}

\newcommand{\alp}{$\alpha$}

@@

\title{The Galactic Chemical Evolution of Carbon}
\author{Daniel A. Boyea}

@@

\begin{document}
% \linespread{1.25}

@@

\begin{titlepage}
   \begin{center}
       \textbf{The Galactic Chemical Evolution of Carbon}\\
       Implications for Stellar Nucleosynthesis\\
       \vspace*{3\baselineskip}
        Undergraduate Research Thesis\\
       \vspace*{3\baselineskip}
    Presented in partial fulfillment of the requirements for graduation \textit{with research distinction in Astronomy and Astrophysics} in the College of Arts and Sciences of The Ohio State University\\
       \vspace*{3\baselineskip}
        by\\
       \vspace*{3\baselineskip}
       {Daniel Alexander Boyea}\\
       \vspace*{3\baselineskip}
       The Ohio State University\\
       April 2023\\
       \vspace*{3\baselineskip}
       Project Advisors\\
       Professor David H. Weinberg, Department of Astronomy \\
       James W. Johnson, Department of Astronomy
       \vfill
   \end{center}
\end{titlepage}

@@

Even though you may have not been able to help me with the science, dear friends and family, you have been no less importaint in the completion of this work. 
Eric, even across the country, you have always been a point of support and stability, and I am always happy to be around you. Anya, you are a wonderful friend and I cannot thank you enough for your unconditional support this year, 
All my friends who are just wonderful, inspiring, and incredible people---Kaia, Maria, Autumn, Simon, Bailee, Aaliyah, Sanskruti, Alyssa, Harrison, Denis, Keith, and so many others. 
And my family (including Arya, my most lovable doodle)---I can not eveen being to thank you enough for simply helping me recover my health and make it to graduation.

@@

\Gce{} aims to understand the chemical and star formation histories of galaxies. At the core of \gce{} models are stellar \gls{yield}s---the amount of each chemical element stars produce (or destroy). Each element is produced in different amounts by different stars, leaving traces of the Galaxy's evolutionary history. Through \textit{Galactic Archeology}, we can reconstruct the history of our galaxy by investigating clues left behind in stars.

Here, I aim to understand C enrichment---where it is produced and how its abundances evolve over time. C is unique nucleosynthetically, one of few light elements (along with N) produced in \agb{} stars \citep[e.g.][]{jennifer19, KL14}. C and N are well-studied elements since they are easy to observe in stellar spectra, even at the lowest metallicities \cite[e.g.][]{fabbian+09, nissen+14, lambert81, laird85, lambert86}. Additionally, C and N are used as age indicators in \gls{rgb} stars \citep{martig16, MG15, hasselquist19, vincenzo+21}.

Gas-phase [C/O] ratios%
%
\footnote{In this paper, I use the standard notation for chemical abundances. $[A/B] = \log_{10}\left(A/B\right) - \log_{10}\left(A_{\sun}/B_{\sun}\right)$, i.e. $[A/B]$ is the logarithm of the ratio between A and B, scales such that $[A/B]=0$ for the sun. Solar abundances are as measured in \citet{asplund+09}}%
%
---observed in very low \gls{metallicity}, high redshift \dla{}---decrease with increasing [O/H] \citep{FN15, cooke+17}. Then, [C/O] increases above $\rm [O/H] \approx -1$ (\citealt{berg+19}; see discussion in section \ref{sec:gas}).
While we know C is produced in \agb{} stars and \cc{}, we still have a limited understanding of the magnitude and metallicity-dependence of each process.

@@

Accurate determination of stellar birth abundances for C and N is challenging. When a star enters the \gls{rgb}, material from the CNO-processed core is mixed with the envelope in first dredge up, enhancing N and depleting C \citep{iben67, vincenzo+21,KL14}. Measurements of these evolved stars'  atmospheres will no longer reflect their birth abundances.  Additionally, gas-phase measurements of C are extremely limited as C lacks strong lines in HII regions \citep{skillman+20}.

To this end, I use a sample of \gls{subgiant} stars from the \apogee{} \citep{apogee17}. According to stellar evolution theory and observations \citep{gilroy89, korn+07, lind+08, souto+18, souto19}, these stars have not yet experienced first dredge up but have well-mixed photospheres. Their photosphoric abundances should thus represent their birth composition.  In Fig.~\ref{fig:subgiants}, I plot this sample, selected by the criteria in Roberts et al. (2023, in prep; see Appendix \ref{sec:jack}). These empirical results contain information on the relative C and Mg yields, which I use later in this work.

\begin{figure}[htp]
    \centering
    \includegraphics{subgiants_mgh.pdf}
    \quad\includegraphics{subgiants_mgfe.pdf}
    \caption[APOGEE Subgiants]{The [C/Mg] ratio against [Mg/H] (top) and [Mg/Fe] (bottom) for the \citetjack~sample of \apogee{} \gls{subgiant}s. On the top, I plot high and low-$\alpha$ stars in blue and orange, using the separation defined in Equation \ref{eq:high_alpha}. On the bottom, I color-code stars according to their [Mg/H] abundance.}
    \label{fig:subgiants}
\end{figure}

@@

Theoretical models of stellar \gls{nucleosynthesis} provide the starting point of this investigation. I focus on three primary nucleosynthetic pathways: \agb{} stars, \cc{}, and \ia. Each process has unique timescales and \gls{yield}s, traceable through the tools of \gce. I compare C, produced in \cc{} and \agb\ stars, to Mg and Fe as tracers of \cc{} and \ia{} espectively.

@@

For an element X and star with mass $M$, the stellar yield $\tilde{y}$ is defined as the net production of X relative to $M$, or
\begin{equation}
    \tilde{y}_{\rm X} = \frac{M_{\rm X,\ ejected} - Z_{0, X} M_{\rm ejected}}{M}   
\end{equation}
where $M_{\rm ejected}$ and $M_{\rm X, ejected}$  are the total ejected mass of the envelope and the element X, respectively. For example, if a 1 $M_\odot$ star has $\tilde{y}_{\rm C} = 10^{-3}$, then the star will add $10^{-3}\ M_\odot$ of C to the interstellar medium. 
Although per-star yields are necessary to compute \agb{} star enrichment rates in \gce{}  models, \gls{imfave} yields are useful in interpreting their predictions. For a yield $y$ from a star of mass M and initial metallicity $Z$, the \gls{imfave} yield is given by 
\begin{equation} \label{eq:imf-yield}
    y_{\rm X}(Z) = 
    \int_{M_{\rm min}}^{M_{\rm max}} 
    \tilde{y}_{\rm X}(M, Z)
    \frac{dN}{dM}  \ dM
\end{equation}
where ${dN}/{dM}$ is the normalized \imf, and $M_{\rm min}$ and $M_{\rm max}$ are the minimum and maximum mass of stars, which I take to be $0.08 M_{\sun}$ and $100 M_{\sun}$ respectively. For \agb{} stars, the yields are truncated above 8 $M_{\odot}$. 
To calculate the \imf-weighted net yields, I use the Versatile Integrator for Chemical Evolution code (\VICE\footnote{\VICE~is available at \url{https://github.com/giganano/VICE}}).

@@

\begin{table}
	\centering
    \caption[Fiducial Model]{Yields for the fiducial model in units of \ssp~birth mass. See section \ref{sec:agb} for the definition of \cxi.}
	\label{tab:fiducial_mod}

@@

\begin{figure}[htp]
    \includegraphics[width=\textwidth]{periodic_table.png}
    \caption{The periodic table of elements. Taken from \cite{jennifer19}.}
\end{figure}

@@

An \agb\ star is a low mass ($\lesssim 8 M_{\sun}$) star during its final phase of evolution.  In an \agb\ star, two competing processes determine the outcome of C production: \gls{tdu} and \gls{hbb}.  Third dredge up accompanies thermal pulses in \agb\ stars. During a \gls{tdu} event, material from the CO core is mixed with material exposed to partial He-shell burning, increasing C and O abundances in the envelope \citep{KL14}. If this envelope is lost during the \agb\ phase, then C yields are enhanced.
\gls{hbb} is the activation of proton capture reactions, and the CNO cycle%
\footnote{The CNO (I) cycle is
\isotope[12]{C}(p, $\gamma$)
\isotope[13]{N}($\beta^+ \nu_e$)
\isotope[13]{C}(p, $\gamma$)
\isotope[14]{N}(p, $\gamma$)
\isotope[15]{O}($\beta^{+}\nu_e$)
\isotope[15]{N}(p, \alp)
\isotope[12]{C}.
 Other minor branches of the CNO cycle are of minor importance to C yields
 \citep{solar-fusion}.
}
at the base of the convective envelope. Because the $^{14}$N(p, $\gamma$) proton capture is the slowest component of the CNO cycle \citep{solar-fusion}, the CNO cycle converts nearly all \isotope[12]{C} into \isotope[14]{N}.
As a result, when both \gls{tdu} and \gls{hbb}, \isotope[12]{C} yields are lowered (see discussion in \citealt{james+23} and \citealt{ventura+13}).

    In this work, I explore four different sets of \agb\ star yield tables from literature, providing well-sampled grids in \gls{metallicity} and mass for use in chemical evolution models. 
\begin{itemize}
    \item \cxi: \citet{cristallo+11, cristallo+15}
    \item \kx: \citet{karakas10}
    \item \vxiii: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item \kxvi: \citet{KL16} \citet{karakas+18}
\end{itemize}
I use \cxi table, amplified by a factor of 2.9, as the fiducial \agb\ yield.

Fig.~\ref{fig:y_agb} compares the \agb\ C yield $\tilde{y}_\text{C}^\text{AGB}(M, Z)$ for each model I consider. 
The yields may be negative if a star ejects material with a lower average C abundance than the material the star was formed from.
Most models agree on the qualitative shape of the net fractional \agb\ C yield---% 
stars with masses between \about{2~M_\odot} and \about{4~M_\odot} have the highest fractional C yields, with the mass of the peak increasing and overall yields decreasing with increasing $Z$.  High mass, high $Z$ stars destroy \isotope[12]{C} because they experience both \gls{tdu} and \gls{hbb}, but the latter is much more efficient.

In Fig.~\ref{fig:agb-ssp}, I plot the total production of C by \agb\ stars in a \ssp{} at an age $t$. 
Effectively, this is equivalent to integrating Eq.~\ref{eq:imf-yield} down to $M_{to}$, the turnoff mass for time $t$.
({In our model, the mass-lifetime relation is
$\log \tau_M = 1.02 - 3.57\log M + 0.90 \left(\log M\right)^2$,
where $\tau_M$ is in Gyr,
from \citealt{larson74}.})
As the mass range $2\lesssim M \lesssim 4$ is most important for C production, about half of C production occurs before \about{1} Gyr, similar to \ia\ Fe. 
\kx{} and \kxvi{} weight C production more heavily towards massive stars resulting in a faster enrichment delay time, whereas the \cxi\ and \vxiii\ models predict a slightly longer timescale of \about{1} Gyr, but little to no C is produced more than 2 Gyr after a star formation event. This is in contrast to Fe whos production lasts up to 10 Gyr after a star formation event. As shown in the right panel of Fig.~\ref{fig:agb-ssp}, with increasing $Z$, C enrichment occurs earlier, and C destruction in low-mass stars leads to a declining C abundance at late times.

\begin{figure}[htp]
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[Low mass C yields]{The net fractional \agb\ C yield  plotted as a function of initial stellar mass $M$ and color-coded according to metallicity. The black dashed line shows $\tilde{y}=0$ for reference. Each panel represents yields from one of four \agb\ models: \cxi, \kx{}, \vxiii{}, \kxvi{} (see section \ref{sec:agb}). }
        \label{fig:y_agb}
\end{figure}
\begin{figure}[htp]
    \centering
    \includegraphics[scale=1]{y_agb_t2.pdf}
